Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example

### 2023-05-22 manuscript/main.qmd
introduction small changes, table

Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@Royer2004; @Burke2018]. Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g. @Liu2009; @Wuchter2004]. Recent advances in the synthesis of such data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system. However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of some regions (e.g. a high proportion of samples from tropical regions). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g. @song2019). However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, that requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modeling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modeling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

Spatial gaps in the record of past climate can also be addressed through the integration of additional data. For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Despite this potential, the integration of geochemical proxy data with other sources of information (e.g. ecological data) is rare due to the absence of a quantitative framework [@Burgener2023].

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. Whilst we primarily do so to demonstrate an application of the model, the EECO represents an interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and a potential analogue for extreme climate warming scenarios [@Burke2018].

>>>>

Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018]. Recent advances in the synthesis of palaeoclimate data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, but a fundamental challenge remains: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g. @Liu2009; @Wuchter2004].  However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of some regions (e.g. a high proportion of samples from tropical regions). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g. @song2019). However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modeling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data. Statistical modeling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

Spatial gaps in the record of past climate can also be addressed through the integration of additional data. For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Despite this potential, the integration of geochemical proxy data with other sources of information (e.g. ecological data) has rarely been realised in a rigorous, quantitative framework [@Burgener2023].

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using downsampled, simulated temperature gradients.

We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. We chose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018]. Our integrative approach allows us to shed new light on the long-standing dispute on the shape of the early Eocene temperature gradient between climate modellers and researchers working with the geological and fossil record [@Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].

::: {#table-1}
![](../figures/Table1.png){width="15cm"}
Table of...
:::
